Prospects of detecting gamma-ray emission from galaxy clusters: cosmic rays and dark matter annihilations

### 2011-05-10 paper.tex
corrections

\begin{abstract}
We study the possibility for detecting gamma-ray emission in galaxy
clusters. We consider 1) cosmic ray (CR) induced pion decay which is
thought to dominate the astrophysical signal from clusters, 2)
different representative benchmark models of supersymmetric dark
matter (DM), and 3) leptophilic models of DM annihilation that
recently have been proposed to explain the positron excess seen by
PAMELA and include a Sommerfeld enhancement of the particle physics
cross section over its standard value. To model DM annihilation, we
consider continuum emission as a result of the hadronization of
annihilating neutralinos, internal bremsstrahlung, and inverse Compton
emission from the cosmic microwave background as well as from a
realistic spatial and spectral distribution of dust and stellar light.
After the galactic center, which hosts many astrophysical sources,
galaxy clusters are the best known target class to detect DM
annihilation. We predict Virgo and Fornax to be the brightest DM
sources and find a particularly low CR induced background for
Fornax. For a given minimum substructure mass as given by the DM
free-streaming scale, one needs the largest DM halo mass to maximize
the substructure boost over the smoothly distributed annihilation
flux: for galaxy clusters as the largest collapsed objects, we find a
boost factor of more than 1000. Since the annihilation flux of
substructures is mostly contributed by the regions around the virial
radius, the resulting surface brightness profiles are almost
flat. This makes it very challenging to detect this flux with imaging
air Cherenkov telescopes since their sensitivity drops linearly with
radius and they typically encounter have $5-10$ linear resolution
elements across a cluster; the gamma-ray satellite Fermi is better
suited. Assuming cold dark matter with a substructure mass
distribution down to an Earth mass and using Fermi upper limits, we
rule out the leptophilic models in their present form in 14 clusters,
and limit the boost from Sommerfeld enhancement in Fornax to be less
than 2 which corresponds to a value in the Milky Way of less than
5. Alternatively, if the Sommerfeld enhancement factor is realized in
Nature, this would imply a limiting substructure mass in Fornax of
$\mlim > 10^2\,\msun$ corresponding to a Milky Way value of $\mlim >
5\times 10^3\,\msun$ which would be a hint for an unknown form of warm
dark matter. We find that Fermi three-year data will be able to
constrain the smallest mass of substructures for supersymmetric
benchmark models or alternatively constrain their cross sections for a
given substructure boost. Although the Perseus and Ophiuchus clusters
are expected to have the largest CR induced flux according to our
semi-analytic model that is derived from cosmological hydrodynamical
cluster simulations, their central active galactic nucleus as well as
the Galactic foreground makes it hard to detect this flux in the Fermi
band. The next to best targets to detect the CR flux are Virgo, Coma,
and Norma -- the Fermi upper limits are closing in on our predictions
and will soon start to constrain the underlying CR physics such as
shock acceleration efficiencies or CR transport properties.
\end{abstract}

\del{
--) the flux from annihilating DM is boosted by about a factor 1000 in
  clusters
--) flat surface brightness profiles with substructures
--) Fornax is a great target for indirect DM studies because the relative
  low flux from CRs and high flux from DM
--) difficult to detect DM without a Sommerfeld boost with Cherenkov
  telescopes since the boost from substructures is extended while the
  sensitivity of Cherenkov telescopes drops linearly with radius
  outside the point spread function (CHECK)
--) the IC upscattered photons from stars and dust are dominated in the
  relevant energy regimes by the continuum emission for our DM
  benchmark models
--) we provide a rather simple model to calculate the spatial and spectral
  distribution of dust and stars in cluster (might be used for
  galaxies? with different spatial profile)
--) we rule out the leptophilic models in present form in 14 clusters,
  and limit the boost from Sommerfeld enhancement in the MW to less
  than 10
--) the tightest limits on DM and CRs from Fermi are in the 1-10 GeV
  regime, where Fornax, M49, and NGC4636 are the most constraining
  "clusters" for the DM and 
... to be written...  if $\mlim \approx 10^{-6}\,\msun$ then the boost
  from SFE is smaller than about a factor 5 in the MW and 2 in Fornax, or
  alternatively if SFE correct then  $\mlim > 10^2\,\msun$.
}

@@

{\bf Anders, insert your flux table for various clusters, and discuss it, DONE}

In this paper, we will investigate in some more detail the potential
of several of the most promising galaxy clusters to yield an
annihilation gamma-ray yield which could be observable with present
and planned gamma-ray detectors. In Table~\ref{tab:flux_tab} we show
the gamma-ray flux from four clusters, which are all targets with a
high gamma-ray yield. We derive the emission from two different DM
models and a hadronic CR model and find that Virgo has the highest
gamma-ray flux which emerge from annihilating DM. If the boost from
Sommerfeld enhancement is small, however, then the gamma-ray emission
induced by CRs is dominating most clusters with the exception of a few
clusters where Fornax had the highest flux from DM. In particular, we
find Fornax ($\mvir=10^{14}\,\msun$) and Virgo
($\mvir=2.1\times10^{14}\,\msun$) \del{\cite{}} as the prime targets
for DM observations, Perseus ($\mvir=7.7\times10^{14}\,\msun$) for CR
induced emission and we additionally decide in favor of the well
studied cluster Coma ($\mvir=1.4\times10^{15}\,\msun$) for
comparison. While studying these four clusters in great detail, we
complement the analysis by calculating the expected gamma-ray fluxes
of a nearby X-ray flux complete sample of galaxy clusters (HIFLUGCS,
\citealt{2002ApJ...567..716R}). For previous work related to DM in
clusters, see, e.g., \cite{2006A&A...455...21C,2009PhRvD..80b3005J}.

\begin{table*}
\begin{minipage}{2.0\columnwidth}
  \caption{Gamma-ray flux from various clusters.}
\begin{tabular}{c  c c c  c  c c c}
\hline
\hline
 Cluster &
\multicolumn{3}{c}{$F_{\gamma}(>100\,\rmn{MeV})$ $[\rmn{ph}\,\rmn{cm}^{-2}\,\rmn{s}^{-1}]$:} & &
\multicolumn{3}{c}{$F_{\gamma}(>100\,\rmn{GeV})$ $[\rmn{ph}\,\rmn{cm}^{-2}\,\rmn{s}^{-1}]$:} \\
         & DM-LP $^{(1)}$ & DM-BM-$\Kp$ $^{(2)}$ & CR-$\pi^0$ $^{(3)}$ 
         & & DM-LP $^{(1)}$ & DM-BM-$\Kp$ $^{(2)}$ & CR-$\pi^0$ $^{(3)}$ \\
 \hline
Fornax               & $4.8\times10^{-8}$ & $1.3\times10^{-10}$ &  $2.2\times10^{-10}$ 
& \,\,\,\,\, & $5.6\times10^{-12}$  & $3.6\times10^{-13}$ & $8.3\times10^{-14}$ \\
Coma                 & $1.3\times10^{-8}$ & $8.3\times10^{-11}$ &  $4.0\times10^{-9}$  
& \,\,\,\,\, & $1.4\times10^{-12}$  & $2.3\times10^{-13}$ & $1.5\times10^{-12}$ \\
Virgo                & $1.2\times10^{-7}$ & $4.0\times10^{-10}$ &  $8.7\times10^{-9}$  
& \,\,\,\,\, & $1.3\times10^{-11}$  & $1.1\times10^{-12}$ & $3.2\times10^{-12}$ \\
Perseus \,\,\,\,\,\, & $1.4\times10^{-8}$ & $7.1\times10^{-11}$ &  $1.4\times10^{-8}$  
& \,\,\,\,\, & $1.4\times10^{-12}$  & $2.0\times10^{-13}$ & $5.2\times10^{-12}$ \\
\hline
\hline
\end{tabular}
\begin{quote}
 Notes: \\ 
 (1) The total gamma-ray flux from leptophilic (LP) DM where both the boost 
 from substructures and Sommerfeld enhancement are included.\\
 (2) The total gamma-ray flux from the supersymmetric $\Kp$ benchmark (BM) 
 model where the boost from substructures is included.\\
 (3) Gamma-ray flux induced by CR protons.
 \label{tab:flux_tab}
  \end{quote}
\end{minipage}
\end{table*}

\del{In this work we define the virial radius $\rvir$ of a halo to be
  the radius at which the mean density within is a factor $\Delta=200$
  times the critical density $\rhoc$ of the universe.}

@@

\subsubsection{Leptophilic models}
\label{sect:LP}
The results of PAMELA \cite{Adriani:2008zr} and Fermi-LAT
\cite{Abdo:2009zk} (and to some extent ATIC
\cite{2008Natur.456..362C}, although the bump seen there at 600 GeV
has not been reproduced by other experiments) has caused intense
activity in the theoretical DM community. At first, there was
skepticism whether such a large signal could be compatible with a
WIMP. However, the possibility of an enhanced annihilation rate due to
DM halo substructure has been realized for a long time
\cite{1993ApJ...411..439S,Bergstrom:1998zs,Moore:1999nt}, although it
is difficult to achieve the boost factor of the order of a few hundred
to a thousand in the solar neighborhood, as substructure according to
numerical simulations is most important in the outer portions of the
halo, due to tidal stripping of subhalos in the inner part. Another
potentially very important effect which may explain the large boost
had been found a few years earlier, namely the Sommerfeld
enhancement. This effect, was computed for electromagnetism by Arnold
Sommerfeld many years ago \cite{sommerfeld} and recently rediscovered
in the quantum field theory of DM
\cite{2005PhRvD..71f3528H,2007NuPhB.787..152C,2009PhRvD..79a5014A}.
In quantum mechanics describing electron scattering and electron
positron annihilation, it is caused by the distortion of the plane
wave describing the relative motion of the annihilating particle pair,
due to the near formation of a bound state caused by photon
exchange. In the ladder approximation for QED, one reproduces the
Sommerfeld effect, and the square of the relative wave function at the
origin (which enters into the probability for the short-distance
process of annihilation) is increased by the factor
\cite{2009PhRvD..79a5014A}
$$
S_{QED}=\frac{|\psi(0)|^2}{|\psi^{(0)}(0)|^2}=
\left|\frac{\frac{\pi\alpha}{v}}{1-e^{-\frac{\pi\alpha}{v}}}\right|,
$$ 
with $\alpha =e^2/\hbar c$ the fine-structure constant, and $v$ the relative
velocity. This amounts to $S_{QED}=\pi\alpha/v$ for small velocities. For a
Yukawa-like particle of mass $m_\phi$, mediating a weak attractive force with
coupling constant $\alpha_Y$ between DM particles $\chi$ with mass
$m_\chi$, the small-velocity limit of the enhancement becomes instead
$$
S_Y\sim\frac{\alpha_Y m_\chi}{m_\phi}.
$$ In the general case, with mediators that may also excite virtual
charged particles in the DM sector, one has to solve numerically a
coupled system of differential equations using appropriate boundary
conditions
\cite{2005PhRvD..71f3528H,2007NuPhB.787..152C,2009PhRvD..79a5014A}. In
some cases, the enhancement factor $S$ can be as high as several
hundred to a few thousand, depending however on the exact parameters
of the theory. The effect is usually strongly velocity-dependent,
depending on velocity as $1/v$ or, in the fine-tuning case of being
very near resonance, as $1/v^2$. This means that in a virialised
system (such as a galaxy cluster) with large velocity dispersion
$v_{\rm cl}$ the Sommerfeld enhancement will be smaller than the one
expected in a single galaxy such as the Milky Way, roughly by a factor
$v_{\rm MW}/v_{\rm cl}$.

@@

To connect easily to our previous work \cite{2009PhRvL.103r1302P}, we
choose as our benchmark leptophilic model one with mass $1.6$~TeV and
Sommerfeld boost $\B_\sfe(\sigv=200\,km/s)\approx 300$ for a Milky Way
halo and scale this appropriately down for cluster halos, using {\bf
  Anders: insert Sommerfeld boost formula as a function of halo mass, done}
\begin{equation}
\B_\sfe(\sigv) = 0.2\, \frac{c}{\sigv} = 
49.4 \left(\frac{\mvir}{10^{15}\,\msun}\right)^{1/3}\,,
\label{eq:B_sfe}
\end{equation}
where the velocity dispersion $\sigv$ for clusters is derived in
\cite{2005RvMP...77..207V}. Clusters have a larger velocity
dispersion and an accordingly smaller Sommerfeld boost factor.

@@

The the wave function of the lightest neutralino can be written
$$
\tilde\chi_1=a_1\tilde B+a_2\tilde W_3+a_3 \tilde H_1^0+a_4\tilde H_2^0,
$$
with 
$$
\sum_{i=1}^4 |a_i|^2=1.
$$
the gaugino fraction of a given neutralino is $|a_1|^2+|a_2|^2$ whereas $|a_3|^2+|a_4|^2$ is the higgsino fraction.

@@

\begin{table}
\begin{tabular}{cccc}
\hline\hline
      BM &  $m_{\chi}$     & $\Omega_{\chi} h^2$ & $\sigmaannv$\\
         &  $[\rmn{GeV}]$ &                    & $[\rmn{cm}^3\,s^{-1}]$\\
\hline
$\Ip$  & 140 & 0.09 & $4.0\times 10^{-27}$ \\
$\Jp$  & 315 & 0.12 & $3.3\times 10^{-28}$ \\
$\Kp$  & 570 & 0.10 & $4.4\times 10^{-26}$ \\
$\Js$  & 234 & 0.09 & $8.9\times 10^{-29}$ \\
\hline\hline
\end{tabular}
 \caption{Relevant parameters for the four benchmark models. The mass
   of the DM particles are given denoted by $m_{\chi}$, the relic
   density by $\Omega_{\chi} h^2$ and the annihilation rate today by
   $\sigmaannv$.\label{tab:BMpara}}
\end{table}
{\bf Anders: add DM particle mass and cross section to all these 4
  models, done, might consider adding the detailed particle physics
  info as well }

@@

The differential photon flux within a given solid angle $\Delta
\Omega$ along a line-of-sight (los) is given by
\begin{equation}
\label{eq:dflux}
\frac{\dd \fg}{\dd \eg} \equiv \frac{\dd^3 \ngamma}{\dd A \,\dd t\, \dd
  \egamma} = \frac{1}{2}\int_{\Delta\Omega} \dd\psi \sin\psi \frac{\dd \sg}{\dd \eg}(\psi,\eg)\,,
\end{equation}
where
\begin{equation}
\label{eq:dtildeflux}
\frac{\dd \sg}{\dd \eg}(\psi,>\eg) = \int_{\Delta\Omega} \dd\Omega \int_{\rm los}
\dd l\, q_{\rm sum}\left(\eg,r\right)\Lambda(\theta)\,,
\end{equation}
and $\sg(\psi, >\eg)$ denotes the surface brightness above the photon
energy $\eg$.  The integration along the line-of-sight $l$, in the
direction $\psi$ that the detector is pointing, is parameterized such
that the radius of the source $r=\sqrt{l^2+D_\clu^2-2 D_\clu
  l\cos\Psi}$, where $D_\clu$ is the distance from the Earth to the
center of the cluster halo and
$\cos\Psi\equiv\cos\theta\cos\psi-\cos\varphi\sin\theta\sin\psi$ {\bf
  Anders: I do not understand this equation: why galactic center?
  changed to earth, doen not really matter}. The angular integration
$\dd \Omega= \sin\theta\dd \theta \,\dd \varphi$ is performed over a
cone centered around $\psi$ and the opening angle $\Delta \Omega$ is
typically taken to be a few times the point spread function (PSF)
$\psf$. The limited angular resolution results in a probability that a
photon coming from a direction $\psi$' is instead reconstructed to a
direction $\psi$, where the underlying probability distribution follow
a Gaussian:
\begin{equation}
\Lambda(\theta)=\frac{1}{2\pi\psf^2}
\,\rmn{exp}\left[-\frac{\theta^2}{2\psf^2}\right]\,,
\quad \rmn{where}\quad \theta=\psi'-\psi \,.
\end{equation}
We denote the total differential source function by $q_{\rm
  sum}(\eg,r)$, where we include contributions from five main
processes; leptophilic DM annihilating either indirectly through
$\mu^+/\mu^-$ to $\rmn{e}^+/\rmn{e}^-$ or directly to
$\rmn{e}^+/\rmn{e}^-$ pairs that IC upscatter background photons
(LP-IC), leptophilic DM emitting final state radiation (LP-FSR),
supersymmetric DM benchmark models where annihilating neutralinos
generate $\rmn{e}^+/\rmn{e}^-$ pairs that upscatter background photons
(BM-IC) and emit a continuum as well as final state radiation
(BM-Cont), and CR proton induced $\pi^0$ that decays into gamma-rays
(CR-$\pi^0$). The source function is given by
\begin{equation}
q_\rmn{sum} (\eg,r) = \qCR(\eg,r)+
\sum_i \,q_{\rmn{sm},i}(\eg,r)\,\B_{\rmn{tot},i}(\sigv,r)
\end{equation}
the differential CR to gamma-ray source function is denoted by
$\qCR(\eg,r)$ (see Sect.~\ref{sect:CRs} and \cite{2010MNRAS.409..449P}
for further details). The subscript $i$ runs over the gamma-ray
producing DM channels and the total differential boost factor for DM
is given by:
\begin{eqnarray}
\B_{\rmn{tot},i}(r,\sigv) = \left\{\begin{array}{cc}
\B_\sfe(\sigv)\,\B_{\sub,i}(r) &\rmn{for\,\,LP}\\
\B_{\sub,i}(r) &\rmn{for\,\,BM\,.}\end{array}\right.
\end{eqnarray}
It is the product of enhancement factors from SFE $\B_\sfe(\sigv)$
(see Eq.~\ref{eq:B_sfe} and Sect.~\ref{sect:LP}) and from substructure
enhancement over the smooth halo contribution $\B_{\sub,i}(r) =
1+\rho_{\sub,i}^2(r)/\rho^2(r)$ (see
Eqs.~\ref{eq:rho_sub}-\ref{eq:xvir} and Sect.~\ref{sect:subst}). {\bf
  Anders: please add equation for $\B_{\sub,i}(M_{halo})$ here and
  reference $B_{sfe}$ equation. added reference to both eqns} The DM
source function from the smooth halo for each process is written in
the form:
\begin{equation}
\label{eq:q_sm}
q_{\rmn{sm},i} (\egamma,r) = \sum_j
\frac{\dd \ngammaj}{\dd E_\gamma} \Gamma_j(r)\,,
\end{equation}
where the annihilation rate density is given by 
\begin{equation}
\label{eq:ann_rate}
\Gamma_j(r) = \frac{1}{2} \left[\frac{\rho(r)}{\mx}\right]^2 
\, \sigmaannv_j\,.
\end{equation}
Here runs the subscript $j$ over all kinematically allowed gamma-ray
producing channels, each with the spectrum $\frac{\dd
  \ngammaj}{\dd\eg}$ and annihilation cross-section $\sigmaannv_j$.
We denote the DM mass with $\mx$ and the smooth DM density profile
with $\rho(r)$.

>>>>

\begin{abstract}
  We study the possibility for detecting gamma-ray emission in galaxy
  clusters. We consider 1) cosmic ray (CR) induced pion decay which is
  thought to dominate the astrophysical signal from clusters, 2)
  different representative benchmark models of supersymmetric dark
  matter (DM), and 3) leptophilic models of DM annihilation that
  recently have been proposed to explain the positron excess seen by
  PAMELA and include a Sommerfeld enhancement of the particle physics
  cross section over its standard value. To model DM annihilation, we
  consider continuum emission as a result of the hadronization of
  annihilating neutralinos, internal bremsstrahlung, and inverse
  Compton emission from the cosmic microwave background as well as
  from a realistic spatial and spectral distribution of dust and
  stellar light.  After the galactic center, which hosts many
  astrophysical sources, galaxy clusters are the best known target
  class to detect DM annihilation. We predict Virgo and Fornax to be
  the brightest DM sources and find a particularly low CR induced
  background for Fornax. For a given minimum substructure mass as
  given by the DM free-streaming scale, one needs the largest DM halo
  mass to maximize the substructure boost over the smoothly
  distributed annihilation flux: for galaxy clusters as the largest
  collapsed objects, we find a boost factor of more than 1000. Since
  the annihilation flux of substructures is mostly contributed by the
  regions around the virial radius, the resulting surface brightness
  profiles are almost flat. This makes it very challenging to detect
  this flux with imaging air Cherenkov telescopes since their
  sensitivity drops linearly with radius and they typically encounter
  have $5-10$ linear resolution elements across a cluster; the
  gamma-ray satellite Fermi is better suited. Assuming cold dark
  matter with a substructure mass distribution down to an Earth mass
  and using Fermi upper limits, we rule out the leptophilic models in
  their present form in 14 clusters, and limit the boost from
  Sommerfeld enhancement in Fornax to be less than 2 which corresponds
  to a value in the Milky Way of less than 5. Alternatively, if the
  Sommerfeld enhancement factor is realized in Nature, this would
  imply a limiting substructure mass in Fornax of $\mlim >
  10^2\,\msun$ which would be a hint for an unknown form of warm dark
  matter. We find that Fermi three-year data will be able to constrain
  the smallest mass of substructures for supersymmetric benchmark
  models or alternatively constrain their cross sections for a given
  substructure boost. Although the Perseus and Ophiuchus clusters are
  expected to have the largest CR induced flux according to our
  analytic model that is derived from cosmological hydrodynamical
  cluster simulations, their central active galactic nucleus as well
  as the Galactic foreground makes it hard to detect this flux in the
  Fermi band. The next to best targets to detect the CR flux are
  Virgo, Coma, and Norma -- the Fermi upper limits are closing in on
  our predictions and will soon start to constrain the underlying CR
  physics such as shock acceleration efficiencies or CR transport
  properties.
\end{abstract}

@@

In this paper, we will investigate in some more detail the potential
of several of the most promising galaxy clusters to yield an
annihilation gamma-ray yield which could be observable with present
and planned gamma-ray detectors. In Table~\ref{tab:flux_tab} we show
the gamma-ray flux from four promising clusters with a high gamma-ray
yield. We derive the emission from two different DM models and a
hadronic CR model and find that Virgo has the highest gamma-ray flux
from annihilating DM. If the boost from Sommerfeld enhancement is
small, however, then the gamma-ray emission induced by CRs is
dominating most clusters with the exception of a few clusters where
Fornax had the highest flux from DM. In particular, we find Fornax
($\mvir=10^{14}\,\msun$) and Virgo ($\mvir=2.1\times10^{14}\,\msun$)
\del{\cite{}} as the prime targets for DM observations, Perseus
($\mvir=7.7\times10^{14}\,\msun$) for CR induced emission and we
additionally decide in favor of the well studied cluster Coma
($\mvir=1.4\times10^{15}\,\msun$) for comparison. While studying these
four clusters in great detail, we complement the analysis by
calculating the expected gamma-ray fluxes of a nearby X-ray flux
complete sample of galaxy clusters from the Rosat all-sky survey
(extended HIghest X-ray FLUx Galaxy Cluster Sample, HIFLUGCS,
\citealt{2002ApJ...567..716R}).\footnote{Note that we have added the
  Virgo cluster to the sample that we nevertheless refer to as
  extended HIFLUGCS catalogue in the following.}  For previous work
related to DM in clusters, see, e.g.,
\cite{2006A&A...455...21C,2009PhRvD..80b3005J}.

\begin{table*}
\begin{minipage}{2.0\columnwidth}
  \caption{Gamma-ray flux from various clusters.}
\begin{tabular}{c  c c c  c  c c c}
\hline
\hline
 Cluster &
\multicolumn{3}{c}{$F_{\gamma}(>100\,\rmn{MeV})$ $[\rmn{ph}\,\rmn{cm}^{-2}\,\rmn{s}^{-1}]$:} & &
\multicolumn{3}{c}{$F_{\gamma}(>100\,\rmn{GeV})$ $[\rmn{ph}\,\rmn{cm}^{-2}\,\rmn{s}^{-1}]$:} \\
         & DM-LP $^{(1)}$ & DM-BM-$\Kp$ $^{(2)}$ & CR-$\pi^0$ $^{(3)}$ 
         & & DM-LP $^{(1)}$ & DM-BM-$\Kp$ $^{(2)}$ & CR-$\pi^0$ $^{(3)}$ \\
 \hline
Fornax               & $4.8\times10^{-8}$ & $1.3\times10^{-10}$ &  $2.2\times10^{-10}$ 
& \,\,\,\,\, & $5.6\times10^{-12}$  & $3.6\times10^{-13}$ & $8.3\times10^{-14}$ \\
Coma                 & $1.3\times10^{-8}$ & $8.3\times10^{-11}$ &  $4.0\times10^{-9}$  
& \,\,\,\,\, & $1.4\times10^{-12}$  & $2.3\times10^{-13}$ & $1.5\times10^{-12}$ \\
Virgo                & $1.2\times10^{-7}$ & $4.0\times10^{-10}$ &  $8.7\times10^{-9}$  
& \,\,\,\,\, & $1.3\times10^{-11}$  & $1.1\times10^{-12}$ & $3.2\times10^{-12}$ \\
Perseus \,\,\,\,\,\, & $1.4\times10^{-8}$ & $7.1\times10^{-11}$ &  $1.4\times10^{-8}$  
& \,\,\,\,\, & $1.4\times10^{-12}$  & $2.0\times10^{-13}$ & $5.2\times10^{-12}$ \\
\hline
\hline
\end{tabular}
\begin{quote}
  Notes: \\
  (1) The total gamma-ray flux from leptophilic (LP) DM where both the
  boost from substructures and Sommerfeld enhancement are
  included. Note that all these values are in conflict with upper limits from
  Fermi (see the following sections for detail).\\
 (2) The total gamma-ray flux from the supersymmetric $\Kp$ benchmark (BM) 
 model where the boost from substructures is included.\\
 (3) Gamma-ray flux induced by CR protons.
 \label{tab:flux_tab}
  \end{quote}
\end{minipage}
\end{table*}

@@

\subsubsection{Leptophilic models}
\label{sect:LP}
The results of PAMELA \cite{Adriani:2008zr} and Fermi-LAT
\cite{Abdo:2009zk} (and to some extent ATIC
\cite{2008Natur.456..362C}, although the bump seen there at 600 GeV
has not been reproduced by other experiments) has caused intense
activity in the theoretical DM community. At first, there was
skepticism whether such a large signal could be compatible with a
WIMP. However, the possibility of an enhanced annihilation rate due to
DM halo substructure has been realized for a long time
\cite{1993ApJ...411..439S,Bergstrom:1998zs,Moore:1999nt}, although it
is difficult to achieve the boost factor of the order of a few hundred
to a thousand in the solar neighborhood, as substructure according to
numerical simulations is most important in the outer portions of the
halo, due to tidal stripping of subhalos in the inner part. Another
potentially very important effect which may explain the large boost
had been found a few years earlier, namely the Sommerfeld
enhancement. This effect, was computed for electromagnetism by Arnold
Sommerfeld many years ago \cite{sommerfeld} and recently rediscovered
in the quantum field theory of DM
\cite{2005PhRvD..71f3528H,2007NuPhB.787..152C,2009PhRvD..79a5014A}.
In quantum mechanics describing electron scattering and electron
positron annihilation, it is caused by the distortion of the plane
wave describing the relative motion of the annihilating particle pair,
due to the near formation of a bound state caused by photon
exchange. In the ladder approximation for QED, one reproduces the
Sommerfeld effect, and the square of the relative wave function at the
origin (which enters into the probability for the short-distance
process of annihilation) is increased by the factor
\cite{2009PhRvD..79a5014A}
\begin{equation}
S_{QED}=\frac{|\psi(0)|^2}{|\psi^{(0)}(0)|^2}=
\left|\frac{\frac{\pi\alpha}{v}}{1-e^{-\frac{\pi\alpha}{v}}}\right|,
\end{equation}
with $\alpha =e^2/\hbar c$ the fine-structure constant, and $v$ the relative
velocity. This amounts to $S_{QED}=\pi\alpha/v$ for small velocities. For a
Yukawa-like particle of mass $m_\phi$, mediating a weak attractive force with
coupling constant $\alpha_Y$ between DM particles $\chi$ with mass
$m_\chi$, the small-velocity limit of the enhancement becomes instead
\begin{equation}
S_Y\sim\frac{\alpha_Y m_\chi}{m_\phi}.
\label{eq:saturation}
\end{equation}
In the general case, with mediators that may also excite virtual
charged particles in the DM sector, one has to solve numerically a
coupled system of differential equations using appropriate boundary
conditions
\cite{2005PhRvD..71f3528H,2007NuPhB.787..152C,2009PhRvD..79a5014A}. In
some cases, the enhancement factor $S$ can be as high as several
hundred to a few thousand, depending however on the exact parameters
of the theory. The effect is usually strongly velocity-dependent,
depending on velocity as $1/v$ or, in the fine-tuning case of being
very near resonance, as $1/v^2$. This means that in a virialised
system (such as a galaxy cluster) with large velocity dispersion
$v_{\rm cl}$ the Sommerfeld enhancement will be smaller than the one
expected in a single galaxy such as the Milky Way, roughly by a factor
$v_{\rm MW}/v_{\rm cl}$.

@@

To connect easily to our previous work \cite{2009PhRvL.103r1302P}, we
choose as our benchmark leptophilic model one with mass $1.6$~TeV and
Sommerfeld boost $\B_\sfe(\sigv=200\,km/s)\approx 300$ for a Milky Way
halo and scale this appropriately down for cluster halos, using
\begin{equation}
\B_\sfe(\sigv) = 0.2\, \frac{c}{\sigv} = 
49.4 \left(\frac{\mvir}{10^{15}\,\msun}\right)^{-1/3}\,,
\label{eq:B_sfe}
\end{equation}
where the velocity dispersion $\sigv$ for clusters is derived in
\cite{2005RvMP...77..207V}. Clusters have a larger velocity dispersion
and an accordingly smaller Sommerfeld boost factor. Note that this
$1/v$-scaling is valid only for $(v/c) \gtrsim (m_\phi/m_\chi)$; at
smaller velocities and outside resonances, the $1/v$-enhancement
saturates at $m_\phi/m_\chi$ \cite{2008PhRvL.101z1301K}.

@@

The the wave function of the lightest neutralino can be written
\begin{equation}
\tilde\chi_1=a_1\tilde B+a_2\tilde W_3+a_3 \tilde H_1^0+a_4\tilde H_2^0,
\end{equation}
with 
\begin{equation}
\sum_{i=1}^4 |a_i|^2=1.
\end{equation}
the gaugino fraction of a given neutralino is $|a_1|^2+|a_2|^2$ whereas $|a_3|^2+|a_4|^2$ is the higgsino fraction.

@@

\begin{table}
\begin{tabular}{cccc}
\hline\hline
      BM &  $m_{\chi}$     & $\Omega_{\chi} h^2$ & $\sigmaannv$\\
         &  $[\rmn{GeV}]$ &                    & $[\rmn{cm}^3\,s^{-1}]$\\
\hline
$\Ip$  & 140 & 0.09 & $4.0\times 10^{-27}$ \\
$\Jp$  & 315 & 0.12 & $3.3\times 10^{-28}$ \\
$\Kp$  & 570 & 0.10 & $4.4\times 10^{-26}$ \\
$\Js$  & 234 & 0.09 & $8.9\times 10^{-29}$ \\
\hline\hline
\end{tabular}
 \caption{Relevant parameters for the four benchmark models. The mass
   of the DM particles are given denoted by $m_{\chi}$, the relic
   density by $\Omega_{\chi} h^2$ and the annihilation rate today by
   $\sigmaannv$.\label{tab:BMpara}}
\end{table}

@@

The differential photon flux within a given solid angle $\Delta
\Omega$ along a line-of-sight (los) is given by
\begin{equation}
\label{eq:dflux}
\frac{\dd \fg}{\dd \eg} \equiv \frac{\dd^3 \ngamma}{\dd A \,\dd t\, \dd
  \egamma} = \frac{1}{2}\int_{\Delta\Omega} \dd\psi \sin\psi \frac{\dd \sg}{\dd \eg}(\psi,\eg)\,,
\end{equation}
where
\begin{equation}
\label{eq:dtildeflux}
\frac{\dd \sg}{\dd \eg}(\psi,>\eg) = \int_{\Delta\Omega} \dd\Omega \int_{\rm los}
\dd l\, q_{\rm sum}\left(\eg,r\right)\Lambda(\theta)\,,
\end{equation}
and $\sg(\psi, >\eg)$ denotes the surface brightness above the photon
energy $\eg$.  The integration along the line-of-sight $l$, in the
direction $\psi$ that the detector is pointing, is parameterized such
that the radius of the source $r=\sqrt{l^2+D_\clu^2-2 D_\clu
  l\cos\Psi}$, where $D_\clu$ is the distance from the Earth to the
center of the cluster halo and
$\cos\Psi\equiv\cos\theta\cos\psi-\cos\varphi\sin\theta\sin\psi$. The
angular integration $\dd \Omega= \sin\theta\dd \theta \,\dd \varphi$
is performed over a cone centered around $\psi$ and the opening angle
$\Delta \Omega$ is typically taken to be a few times the point spread
function (PSF) $\psf$. The limited angular resolution results in a
probability that a photon coming from a direction $\psi$' is instead
reconstructed to a direction $\psi$, where the underlying probability
distribution follow a Gaussian:
\begin{equation}
\Lambda(\theta)=\frac{1}{2\pi\psf^2}
\,\rmn{exp}\left[-\frac{\theta^2}{2\psf^2}\right]\,,
\quad \rmn{where}\quad \theta=\psi'-\psi \,.
\end{equation}
We denote the total differential source function by $q_{\rm
  sum}(\eg,r)$, where we include contributions from five main
processes; leptophilic DM annihilating either indirectly through
$\mu^+/\mu^-$ to $\rmn{e}^+/\rmn{e}^-$ or directly to
$\rmn{e}^+/\rmn{e}^-$ pairs that IC upscatter background photons
(LP-IC), leptophilic DM emitting final state radiation (LP-FSR),
supersymmetric DM benchmark models where annihilating neutralinos
generate $\rmn{e}^+/\rmn{e}^-$ pairs that upscatter background photons
(BM-IC) and emit a continuum as well as final state radiation
(BM-Cont), and CR proton induced $\pi^0$ that decays into gamma-rays
(CR-$\pi^0$). The source function is given by
\begin{equation}
q_\rmn{sum} (\eg,r) = \qCR(\eg,r)+
\sum_i \,q_{\rmn{sm},i}(\eg,r)\,\B_{\rmn{tot},i}(\sigv,r)
\end{equation}
the differential CR to gamma-ray source function is denoted by
$\qCR(\eg,r)$ (see Sect.~\ref{sect:CRs} and \cite{2010MNRAS.409..449P}
for further details). The subscript $i$ runs over the gamma-ray
producing DM channels and the total differential boost factor for DM
is given by:
\begin{eqnarray}
\B_{\rmn{tot},i}(r,\sigv) = \left\{\begin{array}{cc}
\B_\sfe(\sigv)\,\B_{\sub,i}(r) &\rmn{for\,\,LP}\\
\B_{\sub,i}(r) &\rmn{for\,\,BM\,.}\end{array}\right.
\end{eqnarray}
It is the product of enhancement factors from SFE $\B_\sfe(\sigv)$
(see Eq.~\ref{eq:B_sfe} and Sect.~\ref{sect:LP}) and from substructure
enhancement over the smooth halo contribution $\B_{\sub,i}(r) =
1+\rho_{\sub,i}^2(r)/\rho^2(r)$ (see
Eqs.~\ref{eq:rho_sub}-\ref{eq:xvir} and Sect.~\ref{sect:subst}).  The
DM source function from the smooth halo for each process is written in
the form:
\begin{equation}
\label{eq:q_sm}
q_{\rmn{sm},i} (\egamma,r) = \sum_j
\frac{\dd \ngammaj}{\dd E_\gamma} \Gamma_j(r)\,,
\end{equation}
where the annihilation rate density is given by 
\begin{equation}
\label{eq:ann_rate}
\Gamma_j(r) = \frac{1}{2} \left[\frac{\rho(r)}{\mx}\right]^2 
\, \sigmaannv_j\,.
\end{equation}
Here runs the subscript $j$ over all kinematically allowed gamma-ray
producing channels, each with the spectrum $\frac{\dd
  \ngammaj}{\dd\eg}$ and annihilation cross-section $\sigmaannv_j$.
We denote the DM mass with $\mx$ and the smooth DM density profile
with $\rho(r)$.
